A fault tolerant multi-server system for MMOGs

### 2011-07-08 textos/phd-prospectus/prospectus.tex
minor correction
prospectus polished

Generally, each player controls an entity called \emph{avatar}, which is his representation in the virtual environment -- a character who executes the orders given by the player, therefore interfering in the outcome of the game. To provide a consistent view among the different participants, every action performed by each avatar is processed by a server, which calculates the game state resulting from such action. This new state is then broadcast to the players to whom it might be relevant -- again, the state itself does not need to be consistent, but only what the players see.

To cope with the receiving and processing of thousands of actions and the broadcast of their respective resulting states, the traditional approach is to use a large central server -- usually a cluster with a high-speed, low latency Internet connection -- which provides such cpu power and network capabilities \cite{feng2007wnn}. Decentralized alternatives could be used, so that such expensive infrastructure would not be necessary. For example, a fully decentralized system could be used, where each player is a peer responsible for calculating a portion of the game state and for synchronizing it with other peers \cite{rieche2007ppb, hampel2006ppa, elrhalibi2005abm, iimura2004zfg, knutsson2004pps}.

In such peer-to-peer approach, the players would need to agree upon the outcome of the actions, in order for their view to be consistent. To reduce the traffic between peers, the state updates could be sent only to whom they are relevant. To define what is relevant for each player, the virtual environment could be divided into regions \cite{schiele2007rpp}, so that a small peer group would be formed by players who were inside each of these regions. Players whose avatars were in a given region could form a smaller peer-to-peer group -- possibly reducing also the cost of each agreement -- and their own computers would decide the outcome of the players' actions in that area of the virtual environment. Unfortunately, however, the fully decentralized approach would imply a heavy communication cost on the players, specially when running some agreement protocol. This might not only degrade the quality of the game, but also increase the requirements imposed on the players' computational resources, therefore likely reducing their number.

We propose here, then, a system composed of geographically distributed nodes which may act as game servers -- these nodes could be provided, for example, by volunteers or by companies which might have some commercial interest with the game being hosted. Some works, such as \cite{assiotis2006dam,ng2002msa, chertov2006olb, lee2003sdl} also propose the use of a distributed server, but they consider that the nodes are connected through a high speed and low latency network (e.g. a cluster), what makes their solutions only partially applicable in a scenario with highly dynamic and volatile resources, such as wide-area networks formed with volunteer resources. Such networks have some inherent problems: nodes with low availability and dependability, low bandwidth and low processing power compared to current dedicated MMOG servers maintained by large game companies.

The rest of this document is organized as follows: Section \ref{sec:mmogserver} describes the basic features of an MMOG server; Section \ref{sec:distserver} introduces our proposed distributed server model and describes briefly some of its protocols; Section \ref{sec:achievements} lists some of the works we published so far and Section \ref{sec:futurework} presents the research directions to be followed next.

@@

We consider here persistent state real-time games with virtual environments where each player controls an avatar -- a virtual object that represents him -- by issuing commands to it. Through the avatar, the player interacts with the virtual world and with objects present in it, such as avatars from other players. The basic operation of the game network support system must be as follows: when a player connects to the server, it must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to it (usually based on whether those objects may be seen by the player \cite{bezerra2008a3,ahmed2008dai,minson2005aim}). When a player issues a command to the avatar (e.g. move avatar from one place to another, pick up an object, attack another player's avatar etc.), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

@@

Instead of forwarding each command to every server, now only the servers that manage some object in the target set of a command would process it. The problem with such partial state machine replication is that, in order to execute some command, a group might need the previous state of an object managed by some other group. To solve this, a possibilty would be to have the groups requesting the states they need from other groups. Another solution, probably more latency-efficient, would be every group proactively sending the new states of the objects managed by it to other groups which might need them.

Sending new states to other groups proactively based on the possibility that they might need such states, however, may lead to all groups sending to each other the state changes of every object, degenerating to a fully replicated approach. Nevertheless, we can explore the fact that, in games, objects interact only with other objects situated nearby: we can define the regions of the virtual environment in such a way that only objects either in the same or in neighboring regions would be able to interact -- two regions are neighbors if, and only if, they share a border. As commands are executed by avatars, which are simply objects controlled by players, the result will be that, for each pair of objects in the target set of any command, these two objects will be either located in the same or in neighbor regions. This way, only groups managing regions that share a border would need to forward commands to each other\footnote{Although unlikely, it is possible to have a command sequence $C = \{c_1,c_2,\ldots,c_n\}$ where (1) every $c_i$ depends on the state resulting from $c_{i-1}$, (2) not every $c_i$ is processed by the same group and (3) not all of these groups manage neighbor regions. Nevertheless, the servers can detect such dependencies and forward the object states to the servers which need them. In the worst case, every group but one -- there is no dependency cycle -- would be waiting for another group, so the time needed to process a command would be proportional to the number of groups.}.

@@

As the servers are organized in groups, and every command is sent to one or more of such groups, we can use a multicast protocol. Also, as commands seen by different servers must be handled in the same order, we need total order, so that the state machine replication approach can be used. Besides, the commands sent by each player should be executed in the same order in which this player issued them; therefore, the multicast protocol should provide FIFO ordering.

However, the inter-group communication overhead should be minimized, so that the distributed server would be more scalable. Non-genuine multicast primitives \cite{schiper2009gng} require that control messages keep being exchanged between groups, even when there is no application level message to be sent. On the other hand, although genuine multicast algorithms minimize the number of messages exchanged, they impose a greater latecy on their delivery \cite{schiper2008ica}.

@@

The advantage of a quasi-genuine multicast protocol is that it would be a middle ground between genuine and non-genuine protocols, inheriting some advantages from both -- namely, improved scalability and lower message delivery time --, at the cost of restricting which groups can communicate with each other. Although the system can be reconfigured to change the group pairs that are able to exchange messages, such reconfiguration will have some cost, so this multicast protocol is more suitable for applications where this would not be frequent.

This can be used by our consistency management algorithm because every server group only has to forward commands to a specific set of other groups, which are the ones managing neighbor regions, and this is known as soon as the partitioning has been performed. Changes to the partioning should be performed rarely, as it would incur in some players migrating between servers. Therefore, the set of group pairs that can communicate with each other will not change frequently.

To further decrease the message delivery time, an optimistic delivery will also be employed. By assigning timestamps based on the local clocks -- which are supposed to be more or less synchronized -- and by every server waiting long enough for messages to arrive from another servers before performing the delivery, the final total message order will most likely match the timestamp order, so that the messages could be delivered in only one communication step, although that delivery would have to be confirmed by some conservative delivery.

@@

We are currently workings towards the validation of both the quasi-genuine multicast and the game state consistency management protocols, which require both formal proofs regarding correctness and experimental results to demonstrate their performance. Another direction that we intend to follow is to consider not only crash-stop faults, but also Byzantine faults. By tolerating these faults, the server system would also be resistant to cheating, which can also be seen as Byzantine behavior.

>>>>

Generally, each player controls an entity called \emph{avatar}, which is his representation in the virtual environment -- a character who executes the orders given by the player, therefore interfering in the outcome of the game. To provide a consistent view among the different participants, every action performed by each avatar is processed by a server, which calculates the game state resulting from such action. This new state is then broadcast to the players to whom it might be relevant -- most players will need only the portion of the game state that they can see and interact with.

To cope with the receiving and processing of thousands of actions and the broadcasting of their respective resulting states, the traditional approach is to use a large central server -- usually a cluster with a high-speed, low latency Internet connection -- which provides such cpu power and network capabilities \cite{feng2007wnn}. Decentralized alternatives could be used, so that such expensive infrastructure would not be necessary. One example would be a fully decentralized system, where each player is a peer responsible for calculating a portion of the game state and for synchronizing it with other peers \cite{rieche2007ppb, hampel2006ppa, elrhalibi2005abm, iimura2004zfg, knutsson2004pps}.

In such peer-to-peer approach, the players would need to agree upon the outcome of the actions, in order for their view to be consistent. To reduce the traffic between peers, the state updates could be sent only to whom they are relevant. To define what is relevant for each player, the virtual environment could be divided into regions \cite{schiele2007rpp}, so that players whose avatars were in a given region could form a small peer-to-peer group. This way, instead of all peers having to agree upon the outcome of every action, only the peers in a given group would have to agree upon actions which were relevant to that group. Unfortunately, nevertheless, the fully decentralized approach would imply a heavy communication cost on the players, specially when running some agreement protocol. This might not only degrade the quality of the game, but also increase the requirements imposed on the players' computational resources, therefore reducing the number of potential participants.

We propose here, then, a system composed of geographically distributed nodes which may act as game servers -- these nodes could be provided, for example, by volunteers or by companies which might have some commercial interest with the game being hosted. Some works, such as \cite{assiotis2006dam, ng2002msa, chertov2006olb, lee2003sdl}, also propose the use of a distributed server, but they consider that the nodes are connected through a high speed and low latency network (e.g. a cluster), what makes their solutions only partially applicable in a scenario with highly dynamic and volatile resources, such as wide-area networks formed with volunteer resources. Such networks have some inherent problems: nodes with low availability and dependability, low bandwidth and low processing power compared to current dedicated MMOG servers maintained by large game companies.

The rest of this document is organized as follows: Section \ref{sec:mmogserver} describes the basic features of an MMOG server; Section \ref{sec:distserver} introduces our proposed distributed server model and describes briefly some of its protocols; Section \ref{sec:achievements} lists some of the works we have published so far and Section \ref{sec:futurework} presents the research directions to be followed next.

@@

We consider here persistent state real-time games with virtual environments where each player controls an avatar -- a virtual object that represents him -- by issuing commands to it. Through the avatar, the player interacts with the virtual world and with other objects present in it, such as avatars from other players. The basic operation of the game network support system must be as follows: when a player connects to the server, it must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to him (usually based on whether those objects may be seen by the player \cite{bezerra2008a3,ahmed2008dai,minson2005aim}). When a player issues a command to the avatar (e.g. move avatar from one place to another, pick up an object, attack another player's avatar etc.), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

@@

Instead of forwarding each command to every server, now only the servers that manage some object in the target set of a command would process it. The problem with such partial state machine replication is that, in order to execute some command, a group might need the previous state of an object managed by some other group. To solve this, a possibility would be to have the groups requesting the states they need from other groups. Another solution, probably more latency-efficient, would be every group proactively sending the new states of the objects managed by it to other groups which might need them.

Sending new states to other groups proactively based on the possibility that they might need such states, however, may lead to all groups sending to each other the state changes of every object, degenerating to a fully replicated approach. Nevertheless, we can explore the fact that, in games, objects interact only with other objects situated nearby: we can define the regions of the virtual environment in such a way that only objects either in the same or in neighboring regions would be able to interact -- two regions are neighbors if, and only if, they share a border. As commands are executed by avatars, which are simply objects controlled by players, the result will be that, for each pair of objects in the target set of any command, these two objects will be either located in the same or in neighbor regions. This way, only groups managing neighbor regions would ever execute the same command and each group would never need a state from another group which is not its neighbor\footnote{Although unlikely, it is possible to have a command sequence $C = \{c_1,c_2,\ldots,c_n\}$ where (1) every $c_i$ depends on the state resulting from $c_{i-1}$, (2) not every $c_i$ is processed by the same group and (3) not all of these groups manage neighbor regions. Nevertheless, the servers can detect such dependencies and forward the object states to the servers that need them.% In the worst case, every group but one -- there is no dependency cycle -- would be waiting for another group, so the time needed to process a command would be proportional to the number of groups.
}.

@@

As the servers are organized in groups, and every command is sent to one or more of such groups, we can use a multicast protocol. Also, as commands seen by different servers must be handled in the same order, we need total order, so that the state machine replication approach can be used. Besides, the commands sent by each player should be executed in the same order in which they have been issued; therefore, the multicast protocol should provide FIFO ordering.

However, the inter-group communication overhead should be minimized, so that the distributed server would be more scalable. Non-genuine multicast protocols \cite{schiper2009gng} require that control messages keep being exchanged between groups, even when there is no application level message to be sent. On the other hand, although genuine multicast algorithms minimize the number of messages exchanged, they impose a greater latecy on their delivery \cite{schiper2008ica}.

@@

The advantage of a quasi-genuine multicast protocol is that it would be a middle ground between genuine and non-genuine protocols, inheriting some advantages from both -- namely, improved scalability and lower message delivery time --, at the cost of restricting which groups can communicate with each other. Although the system can be reconfigured to change the group pairs that are able to exchange messages, such reconfiguration will have some cost, so this multicast protocol is more suitable for applications where this would not be frequently required.

This can be used by our consistency management algorithm because every server group only has to forward commands and states to a specific set of other groups, which are the ones managing neighbor regions; they are known as soon as the partitioning has been performed. Changes to the partioning should be made rarely, as it would incur in some players migrating between servers. Besides, changing the regions does not necessary change which groups can communicate with each other. Therefore, our quasi-genuine multicast protocol is well suited for the consistency management protocol we are devising.

To further decrease the message delivery time, an optimistic delivery will also be employed. By assigning timestamps based on the local clocks of the servers -- which are supposed to be more or less synchronized -- and by every server waiting long enough for messages to arrive from other servers before performing the delivery, the final total message order will probably match the timestamp order. This way, messages could be delivered optimistically in only one communication step, although this delivery would have to be eventually confirmed by a conservative one.

@@

We are currently workings towards the validation of the quasi-genuine multicast and the game state consistency management protocols, which require both formal proofs regarding correctness and experimental results to demonstrate their performance. Another direction that we intend to follow is to consider not only crash-stop, but also Byzantine faults. By tolerating these faults, the server system would be resistant to many kinds of cheating \cite{yan2005scc}, which can also be seen as Byzantine behavior.
